Backpropagating Gravitational wave events

### 2022-05-20 src/ms.tex
Collaborative edits to discussion.

The correlations between the ZAMS masses are similar to those found by A21 with the majority of the population preferring primary 
and secondary masses between $60--90\,M_{\odot}$. However, we also find some ZAMS masses which extend up to $150\,M_{\odot}$ which 
is the limit imposed by our assumptions. We explore the formation scenarios which lead to successful GW150914-like mergers and those 
which fail to produce GW150914-like mergers below in Figure~\ref{fig:compare_fixed_variable}.

The correlations between $M_{2,\rm{ZAMS}}$ and $f_{\rm acc}$ illustrate the variety of ways which GW150914-like mergers can be produced.
For binaries with $M_{2,\rm{ZAMS}} < 70\,M_{\odot}$, we find that accretion efficiencies of $f_{\rm acc} > 0.5$ 
are preferred. One reason for this is based on the requirement that the total mass in the binary must remain above the 
total mass of GW150914 and strongly non-conservative mass transfer (i.e. $f_{\rm acc} > 0.5$) reduces the total 
mass of the system. A compounding factor is that as mass leaves the binary due to non-conservative mass transfer, the 
evolution of the binary's orbit is less dramatic and leads to wider binaries on average and thus fewer mergers in a Hubble time.
For binaries with $M_{2,\rm{ZAMS}} \sim 80\,M_{\odot}$, $f_{\rm acc}$ is less constrained. This is because of a preference for 
these secondaries to also have primary masses near $ 80\,M_{\odot}$ and thus enter a double-core common envelope evolution in 
which both stars' envelopes are ejected, leaving behind two stripped helium cores in a tight orbit. In this case, $f_{\rm acc}$ 
does not affect the binary evolution and is thus unconstrained. This effect can also be seen in the ZAMS mass and $q_{\rm crit 3}$ 
panels for masses near $80\,M_{\odot}$ because double-core common-envelope evolution is triggered in \texttt{COSMIC} when the
radii of the two stars touch due to the rapid expansion of the primary star upon helium ignition. In this case, the choice for 
Roche-overflow mass transfer to proceed stably (as prescribed by $f_{\rm{acc}}$) or unstably (as prescribed by $\alpha$) using 
$q_{\rm crit, 3}$ is totally irrelevant, and thus unconstrained. Finally, for binaries with $M_{2,\rm{ZAMS}} > 80$, we find that 
$f_{\rm acc}$ is correlated to decrease with increasing mass. This is due to limits on the total mass which must be ejected from 
the binary to produce BH masses that match GW150914. Binaries with both component masses near $150\,M_{\odot}$ must either go 
through a common envelope evolution (in which case $f_{\rm acc}$ is unconstrained), or very non-conservative mass transfer 
(where $f_{\rm acc}$) is low, to produce BBHs with the proper mass.

Finally, we find that $\alpha$ and $f_{\rm acc}$ are largely uncorrelated, though there exists independent trends in each hyper parameter.
This is not totally unexpected since the physical processes which are described by each hyperparameter, i.e. stable Roche-lobe overflow and 
common envelope evolution are two independent channels. Generally, we find that GW150914-like BBHs tend to prefer larger accretion efficiencies 
and have common envelope ejection efficiencies peaking near $\alpha=1$. One shortcoming of our method implementation is the application of a single 
hyper parameter for $f_acc$ and $\alpha$ for each binary, while both the primary and the secondary star could in principle be defined by their own 
hyperparameters that prescribe the outcomes for when each star fills it's Roche lobe. We reserve full treatment of this for future work but note 
that this improvement could reveal correlations between hyperparameters that are not present in our current analysis.
%\kb{We need to describe the interpretation of this as well.}
%\kw{I think we can just hightlight the fact that we have three kinds of correlation, progenitor-hyper, progenitor-hyper, hyper-hyper,
%since the implication is related to figure @Katie made.}
%\kb{Need to describe the formation channels and gaps here.}

@@

% \kw{Talk about implication on single event posterior.}
Once we obtain a set of binaries which successfully map the ZAMS parameters and hyperparameters to BBH merger masses, 
we re-evolve the set of ZAMS parameters with the same physical assumptions as A21, but vary the common envelope efficiency 
to explore how keeping a fixed model which only varies one hyper-parameter contrasts to our results. 
Figure~\ref{fig:compare_fixed_variable} shows the distribution of ZAMS masses for three models (first through third columns), 
each with a different $\alpha$ but the same hyper-parameters as A21, 
as well as the results of our sampling which allows our hyper-parameters to vary (fourth column). We find that holding the accretion 
efficiency, $f_{\rm{acc}}$, and common envelope ejection efficiency, $\alpha$, to fixed values greatly reduces the ZAMS parameter space 
that produces GW150914-like mergers. In contrast, by allowing the hyper-parameters to fully span the model uncertainty, we 
find that there are distinct ZAMS parameters which produce GW150914-like mergers and conversely, ZAMS parameters which 
fail to produce GW150914-like mergers \emph{regardless of our hyper-parameter choice.} For primary ZAMS masses between 
$\sim95-135\,M_{\odot}$ and secondary ZAMS masses between $\sim50-95\,M_{\odot}$, we find that there is no combination of accretion
efficiency and common envelope ejection efficiency which produces merging BBHs that have masses consistent with GW150914's posteriors.
In this region, if BBHs with masses consistent with GW150914's masses are produced through a combination of $\alpha$ and 
$f_{\rm{acc}}$, they do not merge in a Hubble time. Conversely, in the region of the fourth column of Figure~\ref{fig:compare_fixed_variable}
which does produce GW150914-like mergers, the combination of $\alpha$, $f_{\rm{acc}}$, and the ZAMS masses and orbital separations 
balance to produce BHs with the correct masses that merge within a Hubble time.
%\kw{@Katie, can you put a frame around the fourth column highlighting it is this work?}
%\kb{done!}

@@

% Validity paragraph
To check the validity of the recovered posterior, it is essential to check whether it is consistent with the original posterior in the observable space.
We reproject the recovered posterior to the observable space, and compare the KL divergence between the original posterior and the reprojected posterior.
In particular, we compute KL divergence of the samples obtained from root-finding alone and after MCMC sampling.
For GW150914, the KL divergence between the reprojected posterior and the original posterior is \kw{fill later} nats on average if we use the root-finding result alone,
whereas the MCMC result has a higher KL divergence at \kw{fill} nats.
The KL divergence stays stable for reprojecting with different random seeds. 
In fig.\ref{fig:GW150914_reprojection}, we can see the reprojected posterior from both the root-finding process alone and the MCMC result agrees well with the original posterior,
even though the agreement between the MCMC result and the original posterior is not perfect. 
This degradation of the agreement comes from the fact that we are using a KDE to estimate the posterior density in the observable space.
Note that the result from root-finding alone and the result from MCMC agrees well in the evolutionary parameter space.
Compared to the result shown in a previous study, our result covers almost all the posterior of GW150914 in the masses space,
which means our result is more consistent with the observation.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of our method on the population level, we perform the same analysis for all events in GWTC3.
Plotted in figure \ref{fig:GWTC3_f_acc_mass} is the posterior density in the $m_{1,{\rm GW}}-f_{\rm acc}$ space for most of the events in GWTC3.
Each contour represents the $68\%$ credible interval of the posterior density for that particular event.
Although it does not conclusively suggest a dependency between $f_{\rm acc}$ and $m_{1,{\rm GW}}$,
there is a suggestive trend showing $f_{\rm acc}$ could possibly increase as the mass of the progenitor increases.
This means the stable mass transfer (SMT) phase of a light binary is preferably non-conservative, whereas the heavier the events are, the more likely their SMT is conservative.

Note that some events did not pass the KL divergence test we proposed in section \ref{sec:method}, which we do not include in figure \ref{fig:GWTC3_f_acc_mass}.
Low mass events are more subjected to random fluctuation because natal kick is less modulated by fallback in lighter system,
which can lead to a final system either with or without a merger event randomly,
so the reprojected posterior may not agree with the original posterior in the observable space consistently, hence yielding a higher KL divergence on average.
On the other hand, \texttt{COSMIC} struggles to produce events above the pair instability supernova mass cutoff $m_{1,{\rm GW}}$,
so the posterior in the evolutionary parameter space only correspond part of the posterior in the observable space, therefore events beyond the PISN mass gap has higher KL divergence.
Events with more extreme mass ratios are not likely to be recovered by the method. This could be due to our method assuming a single value for $f_{\rm acc}$ and $\alpha$ as discussed 
in Section~\ref{sec:result}.

The point of figure \ref{fig:GWTC3_f_acc_mass} is to show this method could in principal reveal the correlation between progenitor parameters and hyperparameters on a population level.
A careful treatment of all event in the catalog and discussion related the physical implication of figure \ref{fig:GWTC3_f_acc_mass} is beyond the scope of this paper,
therefore we defer a detail study of the physics related to the population of GW events in the future.
Also, despite some events do not pass the KL divergence test, we can still see the correlation in the region where the events are neither affected by randomness nor the PISN mass gap.

@@

We present a new pathway to understand binary evolution with GW events in this paper.
Instead of forward modelling an assumed distribution of initial binary to the observed population,
we find the corresponding evolutionary parameters for each GW events. 
In this work we first showcase the power of the proposed method with GW150914.
We show the joint posterior of the two events' progenitor parameters and hyperparameters.
In particular, to the authors' knowledge, this is the first time the correlation between progenitor parameters and hyperparameters is explored.
Our work is not only a more efficient way to study GW events' progenitors,
but also allows the possibility of measuring the astrophysics related the binary evolution, especially in capturing the correlation between different hyperparameters. 
In future work, we will extend our method to applying hyperparameters to individual binary components in order to better capture correlations between 
hyperparameters and progenitor parameters.

To further illustrate the power of the method, we show the inferred fraction of accretion during SMT depends on the progenitor masses.
The most obvious advantage our method provides is it returns the joint posterior distribution of progenitor parameters and hyperparameters for each event.
This enables a data-driven way to study the distribution of hyperparameters,
i.e. once we have a catalog of events, each has their own set of posterior in the hyper-parameter space,
we can employ common techniques such as hierarchical Bayesian analysis to fit a population model to the distribution of hyperparameters,
without making specific assumptions such as fixing the hyperparameters for all types of event.

% Rate analysis is now just counting, like what's done in phenomenological model, instead of forward modelling.
Another perk of our method is the "rate" calculation is now independent of assumed initial condition such as a particular star formation rate model,
which is a dominant factor of uncertainty for rate prediction in the past.
Once we have pull back the GW event posterior from the observable space to the evolutionary parameter space,
we can apply the same methods we use to account for the rate in the observable space.
In the observable space, the overall rate is basically a normalization that is defined in the local universe i.e. the local merger rate. \kw{cite}
In the evolutionary parameter space, we can instead fit for the production rate of merger events at some redshift.
This also means one can study the progenitor-formation rate in a data-driven manner instead of assuming a particular model.

>>>>

The correlations between the ZAMS masses are similar to those found by A21 with
the majority of the population preferring primary and secondary masses between
$60$--$90\,M_{\odot}$. However, we also find some ZAMS masses which extend up to
$150\,M_{\odot}$ which is the limit imposed by our assumptions. We explore the
formation scenarios which lead to successful GW150914-like mergers and those
which fail to produce GW150914-like mergers below in
Figure~\ref{fig:compare_fixed_variable}.

The correlations between $M_{2,\rm{ZAMS}}$ and $f_{\rm acc}$ illustrate the
variety of ways which GW150914-like mergers can be produced. For binaries with
$M_{2,\rm{ZAMS}} < 70\,M_{\odot}$, we find that accretion efficiencies of
$f_{\rm acc} > 0.5$ are preferred. One reason for this is based on the
requirement that the total mass in the binary must remain above the total mass
of GW150914 and strongly non-conservative mass transfer (i.e. $f_{\rm acc} <
0.5$) reduces the total mass of the system. A compounding factor is that as mass
leaves the binary due to non-conservative mass transfer, the evolution of the
binary's orbit is less dramatic and leads to wider binaries on average and thus
fewer mergers in a Hubble time. For binaries with $M_{2,\rm{ZAMS}} \sim
80\,M_{\odot}$, $f_{\rm acc}$ is less constrained. This is because of a
preference for these secondaries to also have primary masses near $
80\,M_{\odot}$ and thus enter a double-core common envelope evolution in which
both stars' envelopes are ejected, leaving behind two stripped helium cores in a
tight orbit. In this case, $f_{\rm acc}$ does not affect the binary evolution
and is thus unconstrained. This effect can also be seen in the ZAMS mass and
$q_{\rm crit 3}$ panels for masses near $80\,M_{\odot}$ because double-core
common-envelope evolution is triggered in \texttt{COSMIC} when the radii of the
two stars touch due to the rapid expansion of the primary star upon helium
ignition. In this case, the choice for Roche-overflow mass transfer to proceed
stably (as prescribed by $f_{\rm{acc}}$) or unstably (as prescribed by $\alpha$)
using $q_{\rm crit, 3}$ is totally irrelevant, and thus unconstrained. Finally,
for binaries with $M_{2,\rm{ZAMS}} > 80$, we find that $f_{\rm acc}$ is
correlated to decrease with increasing mass. This is due to limits on the total
mass which must be ejected from the binary to produce BH masses that match
GW150914, the reverse situation to $M_{2,\rm{ZAMS}} < 70 \, M_\odot$. Binaries
with both component masses near $150\,M_{\odot}$ must either go through a common
envelope evolution (in which case $f_{\rm acc}$ is unconstrained), or very
non-conservative mass transfer (where $f_{\rm acc}$) is low, to produce BBHs
with the proper mass.

Finally, we find that $\alpha$ and $f_{\rm acc}$ are largely uncorrelated,
though there exist independent trends in each hyper parameter. This is not
totally unexpected since the physical processes which are described by each
hyperparameter, i.e. stable Roche-lobe overflow and common envelope evolution
are two independent channels. Generally, we find that GW150914-like BBHs tend to
prefer larger accretion efficiencies and have common envelope ejection
efficiencies peaking near $\alpha=1$. One shortcoming of our method
implementation is the application of a single hyper parameter for
$f_{\mathrm{acc}}$ and $\alpha$ for each binary, while both the primary and the
secondary star could in principle be defined by their own hyperparameters that
prescribe the outcomes for when each star fills it's Roche lobe. We reserve full
treatment of this for future work but note that this improvement could reveal
correlations between hyperparameters that are not present in our current
analysis.
%\kb{We need to describe the interpretation of this as well.}
%\kw{I think we can just hightlight the fact that we have three kinds of correlation, progenitor-hyper, progenitor-hyper, hyper-hyper,
%since the implication is related to figure @Katie made.}
%\kb{Need to describe the formation channels and gaps here.}

@@

% \kw{Talk about implication on single event posterior.}
Once we obtain a set of binaries which successfully map the ZAMS parameters and
hyperparameters to BBH merger masses, we re-evolve the set of ZAMS parameters
with the same physical assumptions as A21, but vary the common envelope
efficiency to explore how keeping a fixed model which only varies one
hyper-parameter contrasts to our results.
Figure~\ref{fig:compare_fixed_variable} shows the distribution of ZAMS masses
for three models (first through third columns), each with a different $\alpha$
but the same hyper-parameters as A21, as well as the results of our sampling
which allows our hyper-parameters to vary (fourth column). We find that holding
the accretion efficiency, $f_{\rm{acc}}$, and common envelope ejection
efficiency, $\alpha$, to fixed values greatly reduces the ZAMS parameter space
that produces GW150914-like mergers. In contrast, by allowing the
hyper-parameters to fully span the model uncertainty, we find that there are
distinct ZAMS parameters which produce GW150914-like mergers.

Because we can explore the full evolutionary parameter space, we can also
determine that there are ZAMS parameters which fail to produce GW150914-like
mergers \emph{regardless of our hyper-parameter choice.} For primary ZAMS masses
between $\sim95-135\,M_{\odot}$ and secondary ZAMS masses between
$\sim50-95\,M_{\odot}$, we find that there is no combination of accretion
efficiency and common envelope ejection efficiency which produces merging BBHs
that have masses consistent with GW150914's posteriors. In this region, if BBHs
with masses consistent with GW150914's masses are produced through a combination
of $\alpha$ and $f_{\rm{acc}}$, they do not merge in a Hubble time. Conversely,
in the region of the fourth column of Figure~\ref{fig:compare_fixed_variable}
which does produce GW150914-like mergers, the combination of $\alpha$,
$f_{\rm{acc}}$, and the ZAMS masses and orbital separations balance to produce
BHs with the correct masses that merge within a Hubble time.
%\kw{@Katie, can you put a frame around the fourth column highlighting it is this work?}
%\kb{done!}

@@

% Validity paragraph
\wf{What if we made this a subsection of ``methods'', and then just one sentence here about the values of the KL divergence for GW150914?}
To check the validity of the recovered posterior over evolutionary parameters,
it is essential to check whether it is consistent with the original posterior in
the observable space. We reproject the recovered posterior to the observable
space, and compare the KL divergence between the original posterior and the
reprojected posterior. In particular, we compute KL divergence of the samples
obtained from root-finding alone and after MCMC sampling. \wf{Clean up wording.}
For GW150914, the KL divergence between the reprojected posterior and the
original posterior is \kw{fill later} nats on average if we use the root-finding
result alone, whereas the MCMC result has a higher KL divergence at \kw{fill}
nats. The KL divergence stays stable for reprojecting with different random
seeds. In fig.\ref{fig:GW150914_reprojection}, we can see the reprojected
posterior from both the root-finding process alone and the MCMC result agrees
well with the original posterior, even though the agreement between the MCMC
result and the original posterior is not perfect. This degradation of the
agreement comes from the fact that we are using a KDE to estimate the posterior
density in the observable space. Note that the result from root-finding alone
and the result from MCMC agrees well in the evolutionary parameter space.
Compared to the result shown in a previous study, our result covers almost all
the posterior of GW150914 in the masses space, which means our result is more
consistent with the observation.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of our method on the population level, we
perform the same analysis for all events in GWTC3. Plotted in figure
\ref{fig:GWTC3_f_acc_mass} is the posterior density in the $m_{1,{\rm
GW}}-f_{\rm acc}$ space for most of the events in GWTC3. Each contour represents
the $68\%$ credible interval of the posterior density for that particular event.
There is a suggestive trend showing $f_{\rm acc}$ could increase as the mass of
the progenitor increases. Such a trend implies the stable mass transfer (SMT)
phase of a less massive binary would be preferentially non-conservative, with
more conservative mass transfer in more massive binary systems.

Note that some events did not pass the KL divergence test we proposed in section
\ref{sec:method}; we do not include these systems in figure
\ref{fig:GWTC3_f_acc_mass}. Binaries that form low-mass compact objects have
lower amounts of fallback and tend to have correspondingly larger variance in
their random natal kicks. Larger kicks can unbind the progenitor binary, or lead
to wide binaries that do not merge within a Hubble time.  In these cases, the
extra variance during the reprojection can produce a posterior that may not
agree with the original posterior, hence yielding a higher KL divergence.

For high-mass binaries, \texttt{COSMIC} struggles to produce events above the
pair instability supernova mass cutoff $m_{1,{\rm GW}}$, so the posterior in the
evolutionary parameter space only corresponds to part of the posterior in the
observable space below this cutoff, and therefore events beyond the PISN mass
gap also have higher KL divergence.

Events with more extreme mass ratios are hard to produce with \texttt{COSMIC},
and therefore less likely to be accurately recovered by our method. This could
be due to our method assuming a single value for $f_{\rm acc}$ and $\alpha$ as
discussed in Section~\ref{sec:result}.

Merger observations for which \texttt{COSMIC} struggles are likely to be very
informative about formation channels and binary evolution physics, and are
therefore likely worthy of close study.  We anticipate that future work will
follow up these events in detail.

Figure \ref{fig:GWTC3_f_acc_mass} shows that our method could in principle
reveal the correlation between progenitor parameters and hyperparameters on a
population level. A careful treatment of all events in the catalog and
discussion related to the detailed physical implications of figure
\ref{fig:GWTC3_f_acc_mass} is beyond the scope of this paper.  We defer a
detailed study of the physics related to the population of GW events to future
work.

@@

We present a new pathway to understand binary evolution with GW events in this
paper. Instead of forward modelling an assumed distribution of initial binaries
to the observed population, we find the corresponding evolutionary parameters
event-by-event. In this first work, we showcase the power of the proposed method
with an application to GW150914. We show the joint posterior of the event's
progenitor parameters and hyperparameters.  Our work is both a more efficient
way to study GW event progenitors, and also allows the possibility of
constraining the astrophysics related to binary evolution, especially by
capturing the correlation between hyperparameters in different systems.

Our results suggest that the accretion efficiency during stable mass transfer
may depend on the primary black hole mass.  In general, our method returns the
joint posterior distribution of progenitor parameters and hyperparameters for
each event, which enables a data-driven way to study the distribution of
hyperparameters.  That is, once we have a catalog of events, each has their own
posterior in the hyperparameter space, and we can employ well-known techniques
such as hierarchical Bayesian analysis to fit a population model to the
distribution of hyperparameters; this avoids making overly-specific assumptions
like fixing the hyperparameters for all types of event.

% Rate analysis is now just counting, like what's done in phenomenological model, instead of forward modelling.
Working event-by-event as we do permits post-processing application of
 arbitrarily complicated models of star formation and metalliticy evolution by
 re-weighting the samples we obtain \wf{Cite at least Lieke here}. Once we have
 pulled the GW event posterior back from the observable space to the
 evolutionary parameter space, we can apply the same hierarchical inference
 methods currently used to infer the compact binary population to our progenitor
 population.  In particular, we can use our progenitor population to infer the
 formation rate of compact binary population progenitor systems over cosmic time
 without needing to make artifical and simplistic assumptions about the delay
 time distribution or the metallicity-specific star formation rate \wf{again,
 cite Lieke, and also Vitale, Farr, Ng, Rodriguez?}.  Comparing the formation
 rate of progenitors to the star formation rate provides an avenue to check our
 understanding of binary evolution and the relative rate of other formation
 channels.
